CUBE: An \tcb{Information-optimized parallel} Cosmological $N$-body Algorithm

### 2017-11-07 ms_caf/ms_caf.tex
minor changes

\shortauthors{Yu et al.} \shorttitle{CUBE}
\begin{document}
\title{CUBE: An Information-optimized parallel Cosmological $N$-body Algorithm}
\author{
Hao-Ran Yu$^{1,2,3}$,
Ue-Li Pen$^{2,4,5,6}$,
Xin Wang$^{2}$
}

@@

\begin{abstract}
Cosmological large scale structure $N$-body simulations are computation-light, memory-heavy problems in supercomputing. \tcb{This considerable amount of memory is occupied by the phase space information of particles. We show that keeping only a small fraction of the information gives negligible impact on the accuracy of the simulation.} This algorithm is applied in a new parallel, \tcb{information-optimized}, Particle-Mesh based $N$-body code CUBE, where only 6-byte memory is used to store the information of each particle. \tcb{This is accomplished by storing particle's relative phase space coordinates instead of global values, and thus can use a lower accuracy -- in the format of fixed point as light as 1-byte. The remaining information is given by complimentary density and velocity fields (negligible in memory space) and proper ordering of particles (no extra memory).} Our numerical experiments show that this \tcb{information-optimized $N$-body algorithm provides accurate results within the error of the Particle-Mesh algorithm.} This significant lowering of memory-to-computation ratio breaks the bottleneck of scaling up and speeding up large cosmological $N$-body simulations on multi-core and heterogenous computing systems.

@@

When $N$ is large, the brute force pairwise particle-particle (PP) force brings unaffordable $o(N^2)$ computations. However the workload can be alleviated to $o(N\log N)$ by many algorithms. For example, the ``tree'' algorithms like {\tt GADGET} \citep{2001NewA....6...79S,2005MNRAS.364.1105S} and its simulation ``Millennium'' \citep{2005Natur.435..629S,2012MNRAS.426.2046A}, {\tt TPM} \citep{1995ApJS...98..355X} and {\tt GOTPM} \citep{2004NewA....9..111D}, adaptive grid algorithm including {\tt HYDRA} \citep{1995ApJ...452..797C} and {\tt RAMSES} \citep{2010ascl.soft11007T}, as well as mesh-refined codes \citep{1991ApJ...368L..23C} and moving adaptive particle-mesh codes \citep{1995ApJS..100..269P}. Among these, the standard particle-mesh (PM) algorithm \citep{1988csup.book.....H} is most memory and computational efficient if we focus on large cosmological scales. The load balancing problem is minimized as the matter distribution is rather homogeneous, and the speed benefits from the fast Fourier transform (FFT) libraries, such as {\tt FFTW3} \citep{Frigo05thedesign}. {\tt PMFAST} \citep{2005NewA...10..393M} is first codes to use a 2-level PM algorithm, aiming to put PM codes toward speed, memory compactness and scalability. After subsequent developments on {\tt PMFAST}, {\tt CUBEP3M} \citep{2013MNRAS.436..540H} uses cubic spatial decomposition, and adds PP force and many other features.

>>>>

\shortauthors{Yu et al.} \shorttitle{CUBE}
\begin{document}
\title{CUBE: An \tcb{Information-optimized parallel} Cosmological $N$-body Algorithm}
\author{
Hao-Ran Yu$^{1,2,3}$,
Ue-Li Pen$^{2,4,5,6}$,
Xin Wang$^{2}$
}

@@

\begin{abstract}
Cosmological large scale structure $N$-body simulations are computation-light, memory-heavy problems in supercomputing. \tcb{The considerable amount of memory is usually dominated by an inefficient way of storing more than sufficient phase space information of particles.} We present a new \tcb{parallel, information-optimized}, particle-mesh based $N$-body code CUBE, \tcb{in which information- and memory-efficiency is increased by nearly an order of magnitude.} This is accomplished by storing particle's relative phase space coordinates \tcb{instead of global values, and in the format of fixed point as light as 1-byte. The remaining information is given by complimentary density and velocity fields (negligible in memory space) and proper ordering of particles (no extra memory).} Our numerical experiments show that this \tcb{information-optimized $N$-body algorithm provides accurate results within the error of the particle-mesh algorithm.} This significant lowering of memory-to-computation ratio breaks the bottleneck of scaling up and speeding up large cosmological $N$-body simulations on multi-core and heterogenous computing systems.

@@

When $N$ is large, the brute force pairwise particle-particle (PP) force brings unaffordable $o(N^2)$ computations. However the workload can be alleviated to $o(N\log N)$ by many algorithms. For example, the ``tree'' algorithms like {\tt GADGET} \citep{2001NewA....6...79S,2005MNRAS.364.1105S} and its simulation ``Millennium'' \citep{2005Natur.435..629S,2012MNRAS.426.2046A}, {\tt TPM} \citep{1995ApJS...98..355X} and {\tt GOTPM} \citep{2004NewA....9..111D}, adaptive grid algorithm including {\tt HYDRA} \citep{1995ApJ...452..797C} and {\tt RAMSES} \citep{2010ascl.soft11007T}, as well as mesh-refined codes \citep{1991ApJ...368L..23C} and moving adaptive particle-mesh (PM) codes \citep{1995ApJS..100..269P}. Among these, the standard PM algorithm \citep{1988csup.book.....H} is most memory and computational efficient if we focus on large cosmological scales. The load balancing problem is minimized as the matter distribution is rather homogeneous, and the speed benefits from the fast Fourier transform (FFT) libraries, such as {\tt FFTW3} \citep{Frigo05thedesign}. {\tt PMFAST} \citep{2005NewA...10..393M} is first codes to use a 2-level PM algorithm, aiming to put PM codes toward speed, memory compactness and scalability. After subsequent developments on {\tt PMFAST}, {\tt CUBEP3M} \citep{2013MNRAS.436..540H} uses cubic spatial decomposition, and adds PP force and many other features.
